---
title: "Intro to Text Manipulation in R"
output:
  html_document: default
  word_document: default
---
Many data sets have character strings in them. For example, in a file of tweets from Twitter (which are basically just strings of characters), perhaps you want to search for occurrences of a certain word or twitter handle. Or a character variable in a data set might be location with a city and state abbreviation, and you want to extract those observations with location containing "NY."


In this tutorial, you will learn how to manipulate text data using the package `stringr` and how to match patterns using regular expressions. Some of the commands include:


Command | Description
--------|---------------------------------------------------------
`str_sub` | Extract substring from a given start to end position
`str_detect` |Detect presence/absence of first occurrence of substring 
`str_locate` | Give position (start, end) of first occurrence of substring
`str_locate_all`|Give positions of all occurrences of a substring 
`str_replace` | Replace one substring with another



###1. Extracting and locating substrings


We introduce some basic commands from `stringr`.


The `str_sub` command extracts substrings from a string (that is, a sequence of characters) given the starting and ending position. For instance, to extract the characters in the second through fourth position or each string in fruits:


```{r}
library(stringr)
fruits <- c("apple", "pineapple", "Pear", "orange", "peach", "banana")
str_sub(fruits, 2, 4)
```


**Question 1** What are the characters in the first through third position of each string in fruits?


The `str_detect` command checks to see if any instance of a pattern occurs in a string.


```{r}

str_detect(fruits, "p")  #any occurrence of 'p'?
```
Note that pattern matching is case-sensitive.


To locate the position of a pattern within a string, use `str_locate`:


```{r}
str_locate(fruits, "an")
```


Only the fourth and sixth fruits contain "an." In the case of "banana," note that only the first occurrence of "an" is returned.


To find all instances of "an" within each string:


```{r}
str_locate_all(fruits,"an")
```


**Remark** 


The command `str_locate_all` returns a list. 

```{r}
out <- str_locate_all(fruits, "an")
data.class(out)
out[[6]]
unlist(out)
length(unlist(out))/2    #total number of times "an" occurs in vector fruits
```


###2. Regular expressions###


Now suppose we want to detect or locate words that begin with "p" or end in "e," or match a more complex criteria. A **regular expression** is a sequence of characters that define a pattern.

Let's detect strings that begin with either "p" or "P". The metacharacter "^" is used to indicate the beginning of the string, and "[Pp]" is used to indicate "P" or "p".


```{r}
str_detect(fruits, "^[Pp]")
```


Similarly, the metacharacter "$" is used to signify the end of a string.


```{r}
str_detect(fruits, "e$" )   #end in 'e'
```


The following are other metacharacters that have special meanings and so are reserved:

`* \ + $ { }  [ ]   ^ ? .`

For instance, a period matches any single character: 

gr.y matches gray, grey, gr9y, grEy, etc.


and * indicates 0 or more instances of the preceding character:

xy*z matches xz, xyz, xyyz, xyyyz, xyyyyz, etc.  


To detect the letter "a" followed by 0 or more occurrences of "p", type:


```{r}
str_detect(fruits, "ap*")     
```


Compare this to


```{r}
str_detect(fruits, "ap+")
```


The "+" in front of the "p" indicates that we want one or more occurrences of "p." 


Here is a more complex pattern:


```{r}
str_detect(fruits, "^a(.*)e$") 
```


The anchors `^` and `$` are used to indicate we want strings that begin with the letter a and end with e. The `(.*)` indicates that we want to match 0 or more occurrences of any character. In particular, parentheses can be used to group parts of the pattern for readability.


###3 Example


Suppose we want to extract information on 10 digit United States phone numbers from a text file.


```{r}
a1 <- "Home: 507-645-5489"
a2 <- "Cell: 219.917.9871"
a3 <- "My work phone is 507-202-2332"
a4 <- "I don't have a phone"
info <- c(a1, a2, a3, a4)
info
```


We will now extract just the phone numbers from this string.


The area code must start with a 2 or higher so we use brackets again to indicate a range: [2-9]. The next two digits can be between 0 and 9, so we write [0-9]{2}. For the separator, we use [-.] to indicate either a dash or a period. The complete regular expression is given below:


```{r}
phone <- "([2-9][0-9]{2})[-.]([0-9]{3})[-.]([0-9]{4})"
out <- str_detect(info, phone)
out
```


Again, `str_detect` just indicates the presence or absence of the pattern in question.


```{r}
str_extract(info, phone)
```


Let's anonymize the phone-numbers!


```{r}
str_replace(info, phone, "XXX-XXX-XXXX")
```


**Remarks** 

1. As we noted above, certain characters are reserved. If we want to actually reference them in a regular expression, either put them within a bracket, or use a double forward slash.


```{r}
str_locate(info, "[.]")  #find first instance of period
str_locate(info, "\\.")  #same
str_locate(info, ".")    #first instance of any character
```


2. Metacharacters have different meanings within brackets.


```{r}
str_detect(fruits, "^[Pp]")  #starts with 'P' or 'p'
str_detect(fruits, "[^Pp]")  #any character except 'P' or 'p'
str_detect(fruits, "^[^Pp]") #start with any character except 'P' or 'p'
```


3. See the handout regexp.pdf for a summary of regular expressions.


###4. Matching brackets or html tags


In many cases, you may want to match brackets such as [8] or html tags such as `<table>`.


```{r}
out <- c("abc[8]", "abc[9][20]", "abc[9]def[10][7]", "abc[]")
out

```


In order to better understand what regular expressions are matching here, we will replace pieces of the above strings with the character "X".


To replace the left bracket, we write `\\[`. Next we want to match 0 or more occurrences of any character except the right bracket so we need `[^]]*`. Finally, to match the right bracket `\\]`.


```{r}
str_replace_all(out, "\\[([^]]*)\\]", "X")

```


Compare this to


```{r}
str_replace_all(out, "\\[(.*)\\]", "X")

```


In this case, we match the first left bracket (indicated by the `\\[`), followed by 0 or more instances of *any* character (the (.*) portion), which could be a right bracket until the final right bracket `\\]`.


### On your own

1. Create a vector `veggies` containing "carrot", "bean", "peas", "cabbage", "scallion", "asparagus.

+ Find those strings that contain the pattern "ea".

+ Find those strings that end in "s".

+ Find those strings that contain at least two "a"'s.

+ Find those strings that begin with any letter except "c".

+ Find the starting and ending position of the pattern "ca" in each string.

2. The regular expression `"^[Ss](.*)(t+)(.+)(t+)"` matches "scuttlebutt", "Stetson", and "Scattter", but not "Scatter." Why?


### Resources

* [http://www.regular-expressions.info/rlanguage.html](Regular expressions with R)