---
title: "Introduction to Data scraping with R"
output:
  html_document:
    df_print: paged
---
###Scraping Data From the Web###


Data on the web are often presented in tables. For instance, we can see a list of countries by population in 1900 on [Wikipedia](https://en.wikipedia.org/wiki/List_of_countries_by_population_in_1900)


Web pages are written in HTML (Hyper Text Markup Language) which uses **tags** to describe different aspects of document content. For example, a heading in a document is indicated by `<h1>My Title</h1>` whereas a paragraph would be indicated by `<p>A paragraph of content...</p>`. 


In this tutorial, we will learn how to read data from a table on a web page into R. We will need the package `rvest` to get the data from the web page, and the `stringr` package to clean up the data.


```{r, message = FALSE}
library(rvest)
library(stringr)
```


### 1. Reading data into R with `rvest`


To get the population data on  [Wikipedia](https://en.wikipedia.org/wiki/List_of_countries_by_population_in_1900) into R, we use the `read_html` command from the `xml2` package (which is attached when `rvest` is called) to parse the page to obtain an HTML document. 


We then use the `html_nodes` command that extracts all occurrences of the desired tag. We will be interested in scraping data presented in tables, so in the source code, we look for the table tag: `<table> ... </table>`.


Note: some of the `rvest` commands may be slow depending on your Internet connection and the complexity of the web page.


```{r}
popParse <- read_html("https://en.wikipedia.org/wiki/List_of_countries_by_population_in_1900")

str(popParse)
```


The `read_html` command creates an R object, basically a list, that stores information about the web page.


To extract a table from this web page (this may take some time):


```{r}
popNodes <- html_nodes(popParse, "table")
popNodes
```
There are several tables in this document. By inspecting the output of `popNodes`, we make a guess that we want the third table. (In other cases, trial and error may be required.) We select the third table by using double brackets:


```{r}
pop <- html_table(popNodes, header = TRUE, fill = TRUE)[[3]]
str(pop)

```

### 3. Cleaning the data frame


We now have a workable data frame that we can analyze. Notice that even though the first and third columns are numbers, they are classified as "character." For `Rank`, that is because the first observation is the world population and it is not assigned a rank, but rather, the character "-". The `Population` column is also a character because the numbers have commas in them, plus Morocco's population is given as "8,000,000 [2][3]" to indicate some footnotes. We need to convert these columns to be numeric.


In the case of the <tt>Rank</tt> column, we will remove the world population.


We will also simplify the name of the third column to "Population."


```{r}
pop2 <- pop[-1, ]         #remove row 1
head(pop2)

row.names(pop2) <- NULL    #reset row numbers to start at 1

pop2$Rank <- as.numeric(pop2$Rank) #coerce Rank to numeric

names(pop2)[3] <- "Population" #rename 3rd column
```


To remove the commas in the Population numbers, we will use `str_replace_all` from the `stringr` package.


```{r}
pop2$Population <- str_replace_all(pop2$Population, ",", "")
head(pop2)
```


We still have a problem with Morocco, country number 19. Since it is the only country that is a problem, we can fix this individually, but we will also show a more general way to do this in case you encounter more than one instance.


```{r}
pop2$Population[19]
out <- pop2$Population
out[19] <- "8000000"
out <- as.numeric(out)
head(out, 20)

out2 <- str_replace_all(pop2$Population, "\\[[^]]+\\]", "")
out2 <- as.numeric(out2)
head(out2, 20)

identical(out, out2)

pop2$Population <- out2
```


### 4. A movie box office example


The web site [Box Office Mojo](http://www.boxofficemojo.com) gives statistics on box office earnings of movies. In addition to daily earnings, the web site also maintains lists of yearly and all time record holders.


We will look at the movies in the top 100 in all time movie worldwide grosses in box office receipts. In particular, we will scrape the data from [Box Office Mojo: All Time Box Office](http://www.boxofficemojo.com/alltime/world/?pagenum=1). The dollar amounts are in millions of dollars and the years marked with "^" indicate that the movie had multiple releases.


```{r}
movieParse<- read_html("http://www.boxofficemojo.com/alltime/world/?pagenum=1")

movieTables <- html_nodes(movieParse, "table")
head(movieTables)

movies <- html_table(movieTables, header = TRUE, fill = TRUE)[[3]]
str(movies)
```


We will first clean up the column names and then convert the character variables into numeric variables. 


```{r}
names(movies)[5:9] <- c("DomesticDollars", "DomesticPercentage", "OverseasDollars", "OverseasPercentage", "Year")
```


For the grosses, we note that the numbers are either prefaced by a dollar sign or end with a percentage, both of which will need to be removed. We will also have to remove the commas.


```{r}
out <- str_replace_all(movies$Worldwide, "\\$|,", "" )
head(out)

movies$Worldwide <- as.numeric(out)
```


**Exercise** Clean up the remaining variables.


How many images are there on this web page? Images are coded as
`<img source = "NameOfImageFile.jpg">`, so we search for the selector "img".


```{r}
moviesImg <- html_nodes(movieParse, "img")

moviesImg
```
### 4.  A Billboard Top 100 example

The website [billboard.com](http://www.billboard.com)
keeps track of top songs, albums and artists from the music industry.


One page lists the greatest hot women artists. In the source code, here is one artist's listing:
`<div class="chart-list-item  " data-rank="24" data-artist="" data-title="Dionne Warwick" data-has-content="false">`.

(Note that the current number 1 entry, Madonna, is not listed in this format)

```{r}
webParse <- read_html("https://www.billboard.com/charts/greatest-hot-100-women-artists")
str(webParse)

webNodes <- html_nodes(webParse,".chart-list-item  " ) 
webNodes
```

We now need to extract the name of the artist, given in the `data-title` attribute:
```{r}
webNodes2 <- html_attr(webNodes, "data-title")
webNodes2
```

As we noted earlier, Madonna's entry was not listed in the same format as the others, so we will have to add her manually:

```{r}
top50women <- data.frame(top = c("Madonna", webNodes2))
head(top50women, 5)
```

### On Your Own


* The web site [BikeRaceInfo](http://www.bikeraceinfo.com/tdf/tdfstats.html) has a table with data on past winners of the Tour de France. Create a cleaned-up data frame of this data.

* The web site [NY Times Best Sellers: Hardcover Fiction](http://www.nytimes.com/books/best-sellers/hardcover-fiction) contains a list of best-selling fiction books. Scrape the names of these top books. The list of books are tagged via `<h2>title</h2>`.


### Resources

* [HTML Tutorial](www.w3schools.com)