---
title: "Creating Graphs with ggplot2"
output:
  html_document:
    fig_height: 2
    fig_width: 5
---

### 1. Introduction
It is often necessary to create graphs to effectively communicate key patterns within a dataset. While many software packages allow the user to make basic plots, it can be challenging to create plots that are customized to address a specific idea. While there are numerous ways to create graphs, this tutorial will focus on the R package `ggplot2`, created by Hadley Wickham.


There are two key functions that are used in `ggplot2`:

* `qplot()`or *quick plot* is similar to base plotting functions in R and is primarily used to produce quick and easy graphics.
* `ggplot()`the *grammar of graphics plot* is different from other graphics functions because it uses a particular grammar inspired by Leland Wilkinson's landmark book, *The Grammar of Graphics*, that focused on thinking about, reasoning with and communicating with graphics. It enables layering of independent components to create custom graphics.

```{r message = FALSE, warning = FALSE}
# This tutorial will use the following two packages
library(ggplot2)
library(mosaic)
```

**Data**: In this tutorial, we will use the  [AmesHousing](http://www.amstat.org/publications/jse/v19n3/decock.pdf) data, which provides information on the sales of individual residential properties in Ames, Iowa from 2006 to 2010. The data set contains 2930 observations, and a large number of explanatory variables involved in assessing home values. A full description of this dataset can be found [here](http://www.amstat.org/publications/jse/v19n3/Decock/DataDocumentation.txt).

```{r}
# The csv file should be imported into rstudio:
AmesHousing <- read.csv("https://raw.githubusercontent.com/ds4stats/r-tutorials/master/data-viz/data/AmesHousing.csv")
# str(AmesHousing)
```

### 2. The `qplot` function

In this section, we will briefly provide examples of how the `qplot` function (quick plot) can be used to quickly create basic graphs. Run the code below and answer Questions 1)-5).

```{r, eval=FALSE}
# Create a histogram of housing prices
qplot(data=AmesHousing, x=SalePrice, main ="Histogram of Housing Prices in Ames, Iowa")

# Create a scatterplot of above ground living area by sales price
qplot(data=AmesHousing,x=GrLivArea, y=SalePrice)

# Create a scatterplot with log transformed variables, coloring by a third variable
qplot(data=AmesHousing,x=log(GrLivArea),y=log(SalePrice),color=KitchenQual)

# Create distinct scatterplots for each type of kitchen quality and number of fireplaces
qplot(data=AmesHousing,x=GrLivArea,y=SalePrice,facets=KitchenQual~Fireplaces)

# Create a dotplot of sale prices by kitchen quality
qplot(data=AmesHousing,x=KitchenQual,y=SalePrice)

# Create a boxplot of sale prices by kitchen quality
qplot(data=AmesHousing,x=KitchenQual,y=log(SalePrice),geom="boxplot")
```

**Questions**:  

1) In this dataset, how many houses were sold with four fireplaces?
2) What is the `facets` argument used for?
3) Based upon the data documentation, what are the five different levels for kitchen quality?
4) Do these graphs indicate that the quality of a kitchen could be related to the sale price?
5) In the RStudio console, type `?qplot`. Modify the above code to create a barchart (`geom=bar`) to count the number of sales for each level of kitchen quality. What is the deference between the color and fill command?


### 3. The basic structure of the `ggplot` function

All `ggplot` functions must have at least three components:

*  **data frame**: In this activity we will be using the `AmesHousing` data.
*  **geom**: to determine the type of geometric shape used to display the data, such as line, bar, point, or area.
*  **aes**: to determine how variables in the data are mapped to visual properties (aesthetics) of geoms. This can include x position, y position, color, shape, fill, and size.

Thus the simplest code for a graphic made with `ggplot()` would have one of the the following forms: 

* `ggplot(data, aes(x, y)) + geom_line()` or 
* `ggplot(data) + geom_line(aes(x, y))`. 

Note that the two lines of code above would provide identical results. In the first case, the `aes` is set as the default for all `geoms`. In essense, the same `x` and `y` variables are used throughout the entire graphic. However, as graphics get more complex, it is often best to creating local `aes` mappings for each `geom` as shown in the second line of code.

```{r message=FALSE}
# Create a histogram of housing prices
ggplot(data=AmesHousing) + geom_histogram(mapping = aes(SalePrice))
```

In the above code, the terms `data=` and `mapping=` are not required, but are used for clarification. For example, the following code will produce identical results:  
          `ggplot(AmesHousing) + geom_histogram(aes(SalePrice))`.

```{r message=FALSE}
# Create a scatterplot of above ground living area by sales price
ggplot(data=AmesHousing) + geom_point(mapping= aes(x=GrLivArea, y=SalePrice))
```

**Questions:**  

6) Modify the code for the scatter plot above so that the `aes` is not within the `geom`. The resulting graph should look identical to the one above.
7) Create a scatterplot using `ggplot` with **Fireplaces** as the x-axis and **SalePrice** as the y-axis.


### 4. Customizing graphics using the `ggplot` function

In the following code, we layer additional components onto the two graphs (the histogram and scatter plot) shown above.

```{r}
ggplot(data=AmesHousing) +                         
      geom_histogram(mapping = aes(SalePrice/100000), 
          breaks=seq(0, 7, by = 1), col="red", fill="lightblue") + 
      geom_density(mapping = aes(x=SalePrice/100000, y = (..count..)))  +   
      labs(title="Figure 9: Housing Prices in Ames, Iowa (in $100,000)", 
          x="Sale Price of Individual Homes")   
```

**Remarks:**

* The arguments used in `geom_histogram` re-scale the SalePrice, modify the bin size, and change the color/ fill.
* `geom_density` overlays a density curve on top of the histogram.
* Typically density curves and histrograms have very different scales, here we use `y = (..count..)` to modify the density. Alternatively, we could specify `aes(x = SalePrice/100000, y = (..density..))` in the histogram geom.
* The labs() command adds labels to the graph, in this case a title and an x-axis label. A y-axis label can also be added by using `y = " ".


In the code below we create three scatterplots of the log of the above ground living area by the log of sales price

```{r}
ggplot(data=AmesHousing, aes(x=log(GrLivArea), y=log(SalePrice)) ) +      
  geom_point(shape = 3, color = "darkgreen") +                                     
  geom_smooth(method=lm,  color="green") +                  
  labs(title="Figure 10: Housing Prices in Ames, Iowa")

 
ggplot(data=AmesHousing) + 
  geom_point(aes(x=log(GrLivArea), y=log(SalePrice), color=KitchenQual),shape=2, size=2) + 
  geom_smooth(aes(x=log(GrLivArea), y=log(SalePrice), color=KitchenQual), 
          method=loess, size=1) +                        
  labs(title="Figure 11: Housing Prices in Ames, Iowa") 

ggplot(data=AmesHousing) +
  geom_point(mapping = aes(x=log(GrLivArea), y=log(SalePrice), color=KitchenQual)) +
  geom_smooth(mapping = aes(x=log(GrLivArea), y=log(SalePrice), color=KitchenQual), 
      method=lm, se=FALSE, fullrange=TRUE) +                             
  facet_grid(. ~ Fireplaces) +                      
  labs(title="Figure 12: Housing Prices in Ames, Iowa")
```

**Remarks:**

* `geom_point` is used to create a scatterplot. As shown in Figure 10, multiple shapes can be used as points. The [Data Visualization Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) lists several shape options.
* `geom_smooth` adds a fitted line through the data.  
    + `method=lm` specifies a linear regression line. `method=loess` creates a [smoothed local regression curve](https://en.wikipedia.org/wiki/Local_regression).
    + `se=FALSE` removes the shaded 95% confidence regions around each line. The additional option `level` can change the confidence level.
    + `fullrange=TRUE` extends all regression lines to the same length
*  `facet_grid` and `facet_wrap` commands are used to create multiple plots. In Figure 12, we have created separate scatterplots based upon the number of fireplaces.
*  When assigning fixed characteristics, (such as `color`, `shape` or `size`), the commands occur outside the `aes`, as in Figure 10, `color="green"`. When characteristics are dependent on the data, the command should occur within the `aes`, such as in Figure 11 `color=KitchenQual`.


In the above examples, only a few `geoms` are listed. The [ggplot2 website](http://docs.ggplot2.org/current/) lists each `geom` and gives detailed examples of how they are used. 


**Questions:**  

8) Create a histogram of the above ground living area, `GrLivArea`.
9) Create a scatterplot using `YearBuilt` as the explanatory variable and `SalePrice` as the response variable. Include a regression line, a title, and labels for the x and y axes.
10) Modify the scatterplot in Question 9) so that there is still only one regression line, but the points are colored by the overall condition of the home, `OverallCond`. 


```{r eval=FALSE, echo=FALSE}
# SAMPLE  SOLUTIONS
# A histogram of above ground living area 
ggplot(data=AmesHousing) +                         
  geom_histogram(mapping = aes(GrLivArea))

# Create a scatterplot of above ground living area by sales price
ggplot(data=AmesHousing, aes(x=YearBuilt, y=SalePrice)) +      
  geom_point() +
  geom_smooth(method=lm) +                  
  labs(title="Housing Prices in Ames, Iowa", x="Year Built", y = "Sale Price")

# Create scatterplot and regression lines colored by the kitchen quality rating
ggplot(data=AmesHousing, aes(x=YearBuilt, y=SalePrice, color=OverallQual)) + 
  geom_point() +                               
  geom_smooth(method=lm, se=FALSE) +                        
  labs(title="Housing Prices in Ames, Iowa", x="Year Built", y = "Sale Price")

```

### 5. The `mplot` function

The `mosaic` package includes an `mplot` function that involves a helpful pull-down menu for graphic options. 

**Questions:**  

11) In the RStudio Console, type `> mplot(AmesHousing)` and select `2` for a two-variable plot. Select the gear symbol in the top right corner of the graphics window and choose the following items:

* Graphics System: `ggplot2`
* Type of Plot: `boxplot`
* x-variable: `KitchenQual` representing the condition of the sale
* y-variable: `SalePrice`

After selecting these items, click the `Show Expression` to see the `ggplot2` code used to make the boxplot. Now modify the code to include an appropriate title to the plot.

12) Explore the `mplot` function by creating two new graphs that provide information on the `SalePrice` of homes in Ames, Iowa.


### 6. Additional Considerations with R graphics

**Influence of data types on graphics:** If you use the `str` command after reading data into R, you will notice that each variable is assigned one of the following `types`: Character, Numeric (real numbers), Integer, Complex, or Logical (TRUE/FALSE). In particular, the variable **Fireplaces** in considered an integer. In the code below we try to `color` and `fill` a density graph by an integer value. Notice that the color and fill commands appear to be ignored in the graph.
```{r}
# str(AmesHousing)
ggplot(data=AmesHousing) +                   
  geom_density(aes(SalePrice, color = Fireplaces,  fill = Fireplaces))
```

In the following code, we use the `dplyr` package to modify the AmesHousing data; we first restrict the dataset to only houses with less than three fireplaces and then create a new variable, called **Fireplace2**. The `as.factor` command creates a **factor**, wich is a variable that contains a set of numeric codes with character-valued levels. Notice that the `color` and `fill` command now work properly. 

```{r}
# Create a new data frame with only houses with less than 3 fireplaces
AmesHousing2 <- filter(AmesHousing, Fireplaces < 3)
# Create a new variable called Fireplace2
AmesHousing2 <-mutate(AmesHousing2,Fireplace2=as.factor(Fireplaces))
#str(AmesHousing2)

ggplot(data=AmesHousing2) +                 
  geom_density(aes(SalePrice, color = Fireplace2,  fill = Fireplace2), alpha = 0.2)
```

**Customizing graphs:** In addition to using a **data frame**, **geoms**, and **aes**, several additional components can be added to customize each graph, such as: **stats**, **scales**, **themes**, **positions**, **coordinate systems**, **labels**, and **legends**. We will not discuss all of these components here, but the materials in the references section provide detailed explanations. In the code below we provide a few examples on how to customize graphs.

```{r warning =FALSE}
ggplot(AmesHousing2, aes(x = Fireplace2, y = SalePrice, color = PavedDrive)) +
  geom_boxplot(position = position_dodge(width = 1)) +
  coord_flip()+ 
  labs(title="Housing Prices in Ames, Iowa") +
  theme(plot.title = element_text(family = "Trebuchet MS", color = "blue", face="bold", size=12, hjust=0))
```

**Remarks:**

* `position` is used to address geoms that would take the same space on a graph. In the above boxplot, `position_dodge(width = 1)` adds a space between each box. For scatterplots, `position = position_jitter()` puts spaces between overlapping points.
* `theme` is used to change the style of a graph, but does not change the data or geoms. The above code is used to modify only the title in a boxplot. A better approach for beginners is to choose among themes that were created to customize the overall graph. Common examples are `theme_bw()`, `theme_classic()`, `theme_grey()`, and `theme_minimal()`. You can also install the `ggthemes` package for many more options.


**Questions:**  

13) In the density plot above, explain what the `color`, `fill`, and `alpha` commands are used for. Hint: try running the code with and without these commands or use the [Data Visualization Cheat Sheet]( https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf). 

14) In the boxplot, what is done by the code `coord_flip()`?

15) Create a new boxplot, similar to the one above, but use `theme_bw()` instead of the given theme command. Explain how the graph changes.

16) Use the tab completion feature in RStudio (type theme and hit the `Tab` key to see various options) to determine what theme is the default for most graphs in ggplot.


### 7. On your own

For this activity, you should use the `dplyr` package to manipulate the dataset before making any graphics.

#### Option #1 (there is no second option this time, but I encourage you to do the "More Practice" question)

* Restrict the `AmesHousing` data to only sales under normal conditions. In other words, `Condition1 == Norm`
* Create a new variable called `TotalSqFt = GRLivArea  +  TotalBsmtSF` and remove any homes with more than 3000 total square feet.
* Create a new variable, where `No` indicates no fireplaces in the home and `Yes` indicates at least one fireplace in the home.
* Create any other new variables you believe to be informative.
* With this modified data file, create a graphic involving no more than three explanatory variables that best portrays how to predict sales price. For example, Figure 12 uses a linear model of kitchen quality, above ground square footage, and number of fireplaces to predict sale price. Another example is the figure in the **Customizing graphs** section, which uses the fireplaces and paved driveway to predict sale price.
* In addition to submitting your graphic, you should also submit a paragraph explaining what your graph shows and a short justification of the choices you made. Your graphic uses a newly created, derived variable, you should include an explanation of what that variable is.

#### More Practice

The data "economics" is included in the `ggplot2` package (see the documentation [provided here](https://ggplot2.tidyverse.org/reference/economics.html) for descriptions of the variables).  

```{r}
data("economics")
head(economics)
```

For this question I'd like you to create two line graphs using `facet_grid` or `facet_wrap` showing the yearly mean of two different variables. You may choose any two variables from the economics data (other than date) that you find interesting.  Your graphs should display the *yearly mean* of those variables, and be colored according to the president in office that year. The code below provides the necessary presidential data, which you should merge with the economics data.

```{r}
presidents <- data.frame(year = 1967:2015, president = c(rep("Johnson",2), rep("Nixon", 5), rep("Ford", 3), rep("Carter", 4), rep("Reagan", 8), rep("Bush1", 4), rep("Clinton", 8), rep("Bush2", 8), rep("Obama", 7)))
```

### Additional resources
- https://www.youtube.com/watch?v=HeqHMM4ziXA and https://www.youtube.com/watch?v=n8kYa9vu1l8: Two introductory videos on ggplot2 by Roger Peng.

- https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf: Data Visualization with ggplot2 Cheat Sheet 

-  http://docs.ggplot2.org/current/:    A well-documented list of ggplot2 components with descriptions 

-  http://www.statmethods.net/advgraphs/ggplot2.html:   Quick-R introduction to graphics

- http://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf: Formal documentation of the ggplot2 package

- http://www.ceb-institute.org/bbs/wp-content/uploads/2011/09/handout_ggplot2.pdf: A tutorial on ggplot2 by Hadley Wickham.

- http://stackoverflow.com/tags/ggplot2: Stackoverflow, an online community to share information. 

- http://www.cookbook-r.com/Graphs/: R Graphics Cookbook, a text by Winston Chang
http://ggplot2.org/book/ : Sample chapters of Hadley Wickhams text,  ggplot2: Elegant Graphics for Data Analysis





