---
title: "Introduction to Working with Dates in R"
output:
  html_document: default
  word_document: default
---

In many data files, the date or time of day will be an important variable. In this introductory tutorial, we will learn some basics on how R handles dates.


### Using base R to work with dates


There is an international standard ISO 8610 for dates that has the format `yyyy-mm-dd hh:mm:ss`. We can use the base R command `as.Dates()` to convert dates to this standard. The first argument is a vector with the date(s) to be converted, and the second argument gives the format of these dates:


```{r}
as.Date(c("06-28-2015", "11-25-2016"), format = "%m-%d-%Y")

as.Date(c("07/14/17"), format = "%m/%d/%y")
```

Code | Value
-----|------
`%d`   | Day of month as number
`%m`   | Month as number
`%b`   | Month abbreviated
`%B`   | Month full name
`%y`   | Year, two digits
`%Y`   | Year, four digits


Basic operations on date objects include finding the number of days between dates,


```{r}
x <- as.Date("May 12, 2017", format = "%B %d, %Y")
y <- as.Date("March 5, 2017", format = "%B %d, %Y")
x - y
```

and creating a sequence of dates using, for instance, weekly increments.


```{r}
seq(y, length = 6, by = "week")
```

We can also identify the day of the week, or month or quarter on which a date falls.


```{r}
weekdays(y)
months(y)
quarters(y)
```

Date objects are stored internally in R as the number of days since January 1, 1970.


```{r}
as.numeric(x)
x - as.Date("1970-01-01")
```


### 2. The POSIXct class 


In addition to the date, we may also have time information. The `as.POSIXct` command will convert the date and time information to the ISO 8610 standard format.  POSIXct class date objects will specify a timezone also. 


If you only have a date, then the syntax is similar to what we saw above:


```{r}
w <- as.POSIXct(c("05/24/2017", "10/15/2019"), format = "%m/%d/%Y")
w
data.class(w)
diff(w)
```


The timezone in the output will be system dependent.


To add the time information


```{r}
u <- as.POSIXct("05/24/2017 06:13:10", format = "%m/%d/%Y %H:%M:%S")
u
```


You can use the same commands on the POSIXct class that you used for the Date class:


```{r}
weekdays(w)
quarters(w)

seq(u, length = 5, by = "months")
seq(u, length = 5, by = "hours")

w[1]-u
```


If you specify just the time, then `as.POSIXct` will add the current date:


```{r}
as.POSIXct("08:45", format = "%H:%M")
Sys.time()              #current time

as.POSIXct("08:45", format = "%H:%M", tz = "EST")  #specify time zone
```


The list of timezone names is not set in R but is dependent on the user's operating system. Most operating systems recognize the names given here
[http://en.wikipedia.org/wiki/List_of_tz_database_time_zones](http://en.wikipedia.org/wiki/List_of_tz_database_time_zones)


### 3. The lubridate package


The 'lubridate' package by Grolemund and Wickham is just a wrapper for POSIXct commands: that is, `lubridate` commands have a more intuitive syntax for converting to the POSIXct structure.


```{r}
library(lubridate)
```

**Remark**

The message following the `library` command tells us that there is a command called `dates` in `lubridate` which will supersede (mask) the similarly named command `dates` in base R.


The following commands assume the given dates have the format specified by the command name itself:


```{r message = FALSE}
mdy("05/12/2014") #format of input is month/day/year
dmy("21-06-1997") #format of input is day-month-year
mdy_hms("05/12/2014 11:45:10") #format of input is month/day/year hours:minutes:seconds
```


By default, the `mdy_hms` command represents time using UTC, Universal Coordinated Time Zone, which is a standard defined by the International Telecommunications Union Recommendation. Time zones are then expressed as positive or negative offsets from the UTC. 


To find the day of the week of a date, expressed either as a number or the name of the day:


```{r}
wday("05/12/2014 11:45:10")
wday("05/12/2014 11:45:10", label = TRUE)
```


####Rounding dates


```{r}
u
round_date(u, "month")
round_date(u, "hour")
floor_date(u, "hour")
ceiling_date(u, "month")
```

#### Some common lubridate commands

command    | output
-----------|----------
`today`   | Date with no time
`now`      | Date with time
`year`     | Year  
`month`    | Month
`week`     | Week
`yday`     | Day of year (number)
`mday`     | Day of month (number)
`hour`     | Hour
`minute`   | Minute
`second`   | Second
`tz`     | Time zone
`floor_date` | Round down
`ceiling_date`| Round up
`round_date` | Round
